Active Typing: Conservatively Extending a Type System From Within

### 2014-02-28 att-icfp14/att-icfp14.tex
More icfp progress

\title{@$\lambda$: Conservatively Extending a Type System From Within}

@@

%Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language, 
``General-purpose'' languages like ML, on the other hand, expose mechanisms that admit strong embeddings of many more fragments and occupy what is widely seen as a ``sweet spot'' in the design space. Nevertheless, situations continue to arise in both research and practice where introducing a new fragment would be desirable because an abstraction can still only be weakly defined in terms of existing fragments, or a strong embedding, while possible, is impractically verbose or inefficient:
\begin{enumerate}
\item General-purpose abstractions continue to evolve and admit variants. There are many  variations on record types, for example: records with functional record update, specified field ordering (that is, labeled tuples) or prototypic delegation are either awkward to strongly embed or cannot be. Similarly, sum types also admit variants (e.g. various forms of open sums). Even something as seemingly simple as a type-safe \verb|sprintf| operator requires extending a language like Ocaml  \cite{ocaml-printf}.

@@

These sorts of innovations are generally disseminated as distinct \emph{dialects} of an existing general-purpose language, constructed either as a fork, using tools like compiler generators, DSL frame\-works or  language workbenches, or directly within a particular compiler for the language, sometimes activated by a flag or pragma. This, we argue, is quite unsatisfying: a programmer can choose either a dialect supporting an innovative approach to parallel programming or one that builds in support for statically reasoning about units of measure, but there may not be an available dialect supporting both. Forming such a dialect is alarmingly non-trivial, even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that different combinations of individually sound language fragments remain sound when combined. Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from a language   composition operation, so fragment providers have little choice but to leave this to clients.% We will consider this further in our discussion of related work in Sec. \ref{related-work}.

@@

\paragraph{Contributions}
We draw inspiration from this approach to design a mechanism that makes it possible to declare new type and operator constructors and define,  {using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language, from within. If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a form that can be implemented by this protocol, it can be made strongly embeddable in the external language by introducing, in this way, the necessary constructors and logic.

The protocol constrains extensions to maintain type safety and decidability of type checking and type equality. Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong embeddings of a language fragment can be derived assuming a ``closed world'' (that is, in the completely traditional manner where one constructs a complete language and reasons inductively). The semantics guarantees that they will be conserved in the  ``open world'' by enforcing a form of  representation independence at all extension boundaries. Thus, if a strong embedding can be constructed using some combination of extensions, it can be constructed in the presence of arbitrary combinations of other extensions as well, avoiding the problems faced by language dialects and justifying the inclusion of the mechanism directly into the language.

We will begin in Sec. \ref{core} by introducing a minimal calculus with an actively typed semantics, @$\lambda$. The external language builds in only the $\rightarrow$ fragment and the typed internal language is a simple variant of PCF that permits only weak embeddings of even simple fragments like the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ or ML-style $n$-ary products. We implement these as extensions to explain the mechanism and motivate the constraints that the semantics imposes to make the guarantees that we claim. We then examine the semantics in greater detail and state and describe the key points in the proofs of the metatheory in Sec. \ref{theory}. These two sections contain the most fundamental contributions of this work.

@@

While of surprising expressiveness, @$\lambda$ itself is certainly not capable of expressing the full variety of type systems listed above. Indeed, to guarantee safety and conservativity, the calculus is perhaps too conservative. For example, it only admits fragments for which the typing judgement looks essentially like the typing judgement of the $\rightarrow$ fragment; new contexts cannot be introduced. We discuss this and some of the other reasons why a fragment may not be expressible in @$\lambda$ in Sec. \ref{limitations}. These limitations do not appear to be fundamental to the approach, so we also outline in Sec. \ref{limitations} promising directions for future work. We conclude with a discussion of related work in Sec. \ref{related-work}.

@@

The syntax of Core @$\lambda$ is given in Fig. \ref{grammar}. An example of a program defining type and operator constructors that can be used to strongly embed G\"odel's \textbf{T} in @$\lambda$ is given in Fig. \ref{nat}. We will discuss its semantics and how precisely the strong embedding seen being used in this example works as we go on. Natural numbers can, of course, be strongly embedded in existing languages with a similar usage and asymptotic performance profile (up to function call overhead with an abstract type, for example). We will provide more sophisticated examples where this is less feasible later on. %It consists of:
%a type constructor declaration, $\fvar{Nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}). Following the type constructor declaration, we apply $\fvar{Nat}$ with the trivial index, $\tunit$, to form the type $\tvar{nat}$. Finally, we write an external term that uses the operators associated with $\fvar{nat}$ and the built-in constructor $\fvar{Parr}$, governing partial functions, to define an addition function and compute the addition of the natural numbers  two and two. We will introduce a more convenient concrete syntax in later portions of this thesis; for now we will restrict ourselves to the abstract syntax so that this example can directly aid in understanding the semantics.

@@

A program, $\rho$, consists of a series of static declarations (in the core, only constructor declarations) followed by an external term, $e$. The syntax for external terms contains three core forms: variables, $\lambda$, and a form for all other operator invocations, which we will discuss below. Compiling a program consists of first \emph{kind checking} its static declarations and type-level terms (Fig. \ref{kindprog}, discussed below) then \emph{type checking} the external term and, simultaneously, \emph{translating} it to a term, $\iota$, in the \emph{typed internal language} (Fig. \ref{att}). The key judgement is the \emph{active typing judgement}, relating an external term to a {type} and an internal term, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$: 

@@

This judgement follows standard conventions for the typing judgement of a simply-typed lambda calculus in most respects. The typing context $\Gamma$ maps variables to types and obeys standard structural properties. The key novelties here are 1) that the available type and operator constructors are not fixed by the language, but rather are tracked by the constructor context, $\fvalCtx$, and 2) that a translation is simultaneously derived. There is no separate operational semantics for the external language; the dynamic semantics are defined by this translation to the internal language (which might have a more conventional operational semantics). \todo{cite Harper-Stone / denotational semantics} The two contexts and the external term can be thought of as input to the judgement, while the type and translation are output (we do not consider a bidirectional approach \cite{LovasPfenning}\todo{cite} in this work, though this is likely a promising avenue for future research).

In our calculus, the internal language (IL), consisting of internal terms $\iota$ and internal types $\sigma$, contains partial functions (via the generic fixpoint operator of Plotkin's PCF \cite{PCF}), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines). In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been determined (e.g. perhaps Standard ML or CakeML \cite{cakeml}, leaving translation to a simpler IL to later stages of the compiler).

@@

\inferrule[hide-trans]{
	\fvar{Tycon} \notin \Xi
}{
	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}^{\checkmark}}}{\trepof{\ttype{Tycon}{\tauidx}}}
}
\end{mathpar}
\caption{\small Abstracted internal typing}
\label{ait}
\end{figure}
@$\lambda$ is an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation. More specifically, type-level terms, $\tau$, themselves form a typed lambda calculus. The classifiers of terms at this level are called \emph{kinds}, $\kappa$ to distinguish them from  \emph{types}, which are just one kind  of type-level value. We will write the kind of types as $\kTypeBlur$, though it is also written $\star$, \verb|T| or \verb|Type| in various similarly structured languages (see Sec. \ref{related-work}). The kinding rules for type-level terms are given in Fig. \ref{todo}\todo{refer}.% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions. In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.

In most languages, types are formed by applying one of a fixed collection of \emph{type constructors} to zero or more \emph{indices}. In @$\lambda$, the situation is different. User-defined type constructors are declared at the top-level of a program (or lifted there, in a practical implementation) using \textsf{tycon} (Fig. \ref{kindprog}). Each type constructor in the program must have a unique name, written e.g. \fvar{Nat}. We assume naming conflicts can be avoided by some extrinsic mechanism (e.g. a URI-based scheme). A type constructor must also declare an \emph{index kind}, $\kappaidx$. Types themselves -- type-level values of kind $\kTypeBlur$ -- are introduced by applying a previously declared type constructor to a type-level term of its index kind, $\ttype{Tycon}{\tauidx}$. The kind $\kTypeBlur$ also has an elimination form: a type can be case analyzed against a type constructor in scope to extract its index. We are, to a first approximation, treating type constructors as constructors of a built-in open datatype at the type-level. Like open datatypes, there is no longer a notion of exhaustiveness and the default case is required for case analysis.

 To permit the implementation of interesting type systems, the type-level language includes several kinds other than $\kTypeBlur$. We lift several  standard functional constructs to the type level: unit ($\kunit$), binary sums ($\ksum{\kappa_1}{\kappa_2}$), binary products ($\kpair{\kappa_1}{\kappa_2}$), lists ($\klist{\kappa}$) and integers ($\dint$). We also include labels ($\klabel$), written in a slanted font, e.g. $\tlabel{myLabel}$, which are string-like values that only admit comparisons, here only for equality, and play a distinguished role in the expanded syntax, as we will later discuss. Our first example, $\fvar{Nat}$, is indexed trivially, i.e. by unit kind, $\kunit$, because there is only one natural number type, $\ttype{Nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work. For example, $\fvar{Tuple}$ has index kind $\klist{\kTypeBlur}$ and $\fvar{LabeledTuple}$ has index kind $\klist{\kpair{\klabel}{\kTypeBlur}}$. The type constructor $\fvar{Arrow}$ is included in the initial constructor context and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.

Two type-level terms of kind $\kTypeBlur$ are equivalent if they apply the same constructor, identified by name, to equivalent indices. Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after normalization by imposing the restriction that type constructors can only be indexed by kinds for which equivalence can be decided in this way. All combinations of kinds introduced thus far  have this property and this is sufficient for many interesting examples. We discuss introducing a richer notion of  equivalence of types that  fragment providers can extend as future work in Sec. \ref{limitations}\todo{do this}.% as maintaining the metatheoretic guarantee that typing respects type equivalence would impose a substantial burden in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).

Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.  \cite{Watkins08}\todo{cite}; Ch. 22 of \emph{PFPL} describes a related system \cite{pfpl}). The type-level language does, however, include total functions of conventional arrow kind, written $\karrow{\kappa_1}{\kappa_2}$. Type formation can be wrapped in a type-level function to emulate a first-class type constructor (indeed, such a wrapper could be generated automatically, though we avoid this for simplicity in our semantics). Equivalence at arrow kind does not coincide with $\beta$-equivalence, so type-level functions cannot appear in type indices. Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML. An equality type is defined by the judgement \todo{eq type discussion}. 
Every type constructor also defines a \emph{representation schema} using \textsf{schema}, a type-level function that we will discuss after introducing operators.

\subsection{Operators}
For reasons that we will discuss, our calculus associates every operator  constructor\footnote{We distinguish \emph{term constructors} from \emph{type constructors} and \emph{operator constructors}. There are term constructors at all levels in the calculus. For example, type constructor application is the sole type-forming type-level term constructor. An abstract syntax at the type level like the one used by Harper \cite{pfpl} might write it $\texttt{tcapp}[\fvar{Tycon}](\tau)$, emphasizing this point.} with a type constructor.

User-defined operator constructors are declared using \textsf{opcon} (Fig. \ref{kindprog}). The \emph{fully-qualified name} of every operator constructor, e.g. $\fvar{Nat}.\opvar{z}$, must be unique. Operator constructors are indexed by type-level values, like type constructors, and so also declare an index kind $\kappaidx$. In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but later examples will use more interesting indices. For example, the projection operator for labeled tuples, $\fvar{LabeledTuple}.\opvar{prj}$, has index kind $\klabel$.
Note that neither operator constructors nor operators are first-class type-level values and we do not need to impose any restrictions on their index kinds.

In the external language, an operator is selected and invoked by applying an operator constructor to a type-level index and $n \geq 0$ \emph{arguments}, written $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$. For example, on line 18 of Fig. \ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being invoked to compute two (the syntactic overhead will be relieved later). %\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}. With the ability to ``open'' a type's operators into the context, we could shorten this still to \texttt{s(s(z))}. Alternatively, with the ability to define a TSL in a manner similar to that in Sec. \ref{aparsing}, we might instead just write \texttt{2}.}
To derive the active typing judgement for this form, the semantics invokes  
 the operator constructor's  \emph{definition}: a type-level function that must examine the provided operator index and the recursively determined \emph{derivates} of the arguments to determine a derivate for the operation as a whole,  or indicate an error. A derivate is a type-level representation of the result of deriving the active typing judgement\footnote{Technically, the abstract active typing judgement, which is similar in form and we will introduce shortly.}: an internal term, called the {translation},  paired with a type. Derivates have kind $\kDen$ and introductory form $\tden{\tautrans}{\tautype}$. The elimination form $\tderbind{x}{t}{\tau}{\tau'}$ extracts the translation and type from the derivate $\tau$, binding it to $\tvar{x}$ and $\tvar{t}$ respectively in $\tau'$. 
%Indeed, one might be tempted to simply define derivates as products. However, we treat translations with much care throughout the calculus, so this is not quite the case (we will return to this shortly). 
Because constructing a derivate is not always possible (e.g. when there is a type error in the client's code, or an invalid index was provided), the return kind of the function is an ``option kind'', $\ksum{\kDen}{\kunit}$, where the trivial case indicates a statically-detected error in the client's program. In practice, it would instead require providers to report information about the precise location of the error and an appropriate error message.

>>>>

\title{Active Typing: Conservatively Extending a Type System From Within}

@@

%Having established why a minimal language like $\mathcal{L}\{\rightarrow\forall\}$, while suitable as an internal language, needs to be extended with additional fragments before it is suitable for use as a human-facing external language, 
Modern ``general-purpose'' languages like ML, on the other hand, expose constructs, like datatypes and abstract types, that admit strong embeddings of many simpler fragments, occupying what is widely seen as a ``sweet spot'' in the design space. However, ``general-purpose'' and ''all-purpose'' are quite distinct, and situations continue to arise in both research and practice where introducing a new fragment would be desirable because it can still only be weakly defined in terms of existing fragments, or a strong embedding, while possible, is impractically verbose or inefficient:
\begin{enumerate}
\item General-purpose abstractions continue to evolve and admit variants. There are many  variations on record types, for example: records with functional record update, specified field ordering (that is, labeled tuples) or prototypic delegation are either awkward to strongly embed or cannot be. Similarly, sum types also admit variants (e.g. various forms of open sums). Even something as seemingly simple as a type-safe \verb|sprintf| operator requires extending a language like Ocaml  \cite{ocaml-printf}.

@@

These sorts of innovations are generally disseminated as distinct \emph{dialects} of an existing general-purpose language, constructed either as a fork, using tools like compiler generators, DSL frame\-works or  language workbenches, or directly within a particular compiler for the language, sometimes activated by a flag or pragma. This, we argue, is quite unsatisfying: a programmer can choose either a dialect supporting an innovative approach to parallel programming or one that builds in support for statically reasoning about units of measure, but there may not be an available dialect supporting both. Forming such a dialect is alarmingly non-trivial, even in the rare situation where a common framework has been used or the dialects are implemented within the same compiler, as these mechanisms do not guarantee that different combinations of individually sound language fragments remain sound when combined. Metatheoretic and compiler correctness results can only be derived for the dialect \emph{resulting} from a language   composition operation, so in a diverse fragment ecosystem, fragment providers have little choice but to leave this to clients. Avoiding language composition is difficult, as interoperability between components of an application written in different dialects faces precisely the problems described in item 3.% We will consider this further in our discussion of related work in Sec. \ref{related-work}.

@@

\paragraph{Contributions}
We draw inspiration from this approach to design a mechanism that makes it possible to declare new type and operator constructors and define, \emph{using type-level functions}, their static and dynamic semantics, the latter by translation to a fixed typed internal language. If a language fragment can be weakly defined in terms of the typed internal language and its statics are of a form that can be implemented by this protocol, it can be made strongly embeddable in the external language by introducing, in this way, the necessary constructors and static logic.

The semantics imposes checks on extensions to guarantee type safety and decidability of typing and type equality. Moreover, an important class of lemmas that play a key role in  proofs that an extension enables strong embeddings of a language fragment can be derived assuming a ``closed world'' (that is, in the completely traditional manner where one constructs a complete language and reasons inductively). The semantics guarantees that they will be \emph{conserved} in the  ``open world'' by enforcing a form of  representation independence at all extension boundaries. This avoids the biggest problems of composing language dialects and justifies the inclusion of the mechanism directly into the language.

We will begin in Sec. \ref{core} by introducing a minimal calculus with this kind of \emph{actively typed semantics}, @$\lambda$. The external language begins with only the $\rightarrow$ fragment and the typed internal language is a simple variant of PCF that permits only weak embeddings of even fragments like the $\mathtt{nat}$ fragment of G\"odel's $\mathbf{T}$ or ML-style $n$-ary products. We implement these as extensions to explain the mechanism and motivate the constraints that the semantics imposes. We then examine the semantics in greater detail and state and sketch the key points in the proofs of the metatheory in Sec. \ref{theory}. These are the fundamental contributions of this work.

@@

While of surprising expressiveness, @$\lambda$ itself is certainly not capable of expressing the full variety of type systems listed earlier. Indeed, to guarantee safety and conservativity, our calculus assumes perhaps too much about the type system being encoded. For example, it only admits fragments for which the typing judgement looks essentially like the typing judgement of the $\rightarrow$ fragment; new contexts cannot be introduced. We discuss this and some of the other reasons why a fragment may not be expressible in @$\lambda$ in Sec. \ref{limitations}. Fortunately, these limitations do not appear fundamental to the approach, so we also outline in Sec. \ref{limitations} directions for future work. We conclude with a discussion of related work in Sec. \ref{related-work}.

@@

The syntax of Core @$\lambda$ is given in Fig. \ref{grammar}. An example of a program defining type and operator constructors that can be used to strongly embed G\"odel's \textbf{T} in @$\lambda$ is given in Fig. \ref{nat}. We will discuss its semantics and how precisely the strong embedding, seen being used starting on line 16, works as we go on. Natural numbers can, of course, be strongly embedded in existing languages with a similar usage and asymptotic performance profile (up to function call overhead with an abstract type, for example). We will provide more sophisticated examples where this is less feasible later on. %It consists of:
%a type constructor declaration, $\fvar{Nat}$, indexed trivially, together with three operator constructors, also all indexed trivially, that implement the standard introductory forms for natural numbers as well as the recursor operator (as in G\"odel's T \cite{pfpl}). Following the type constructor declaration, we apply $\fvar{Nat}$ with the trivial index, $\tunit$, to form the type $\tvar{nat}$. Finally, we write an external term that uses the operators associated with $\fvar{nat}$ and the built-in constructor $\fvar{Parr}$, governing partial functions, to define an addition function and compute the addition of the natural numbers  two and two. We will introduce a more convenient concrete syntax in later portions of this thesis; for now we will restrict ourselves to the abstract syntax so that this example can directly aid in understanding the semantics.

@@

A \emph{program}, $\rho$, consists of a series of static declarations (in the core, only constructor declarations) followed by an external term, $e$. The syntax for external terms contains three term formers: variables, $\lambda$, and a form for all other operator invocations, which we will discuss below. Compiling a program consists of first \emph{kind checking} its static declarations and type-level terms (Fig. \ref{kindprog}, discussed below) then \emph{type checking} the external term and, simultaneously, \emph{translating} it to a term, $\iota$, in the {typed internal language} (Fig. \ref{att}). The key judgement is the \emph{active typing judgement}, relating an external term to a {type} and an internal term, called its \emph{translation}, under \emph{typing context} $\Gamma$ and \emph{constructor context} $\Phi$: 

@@

This judgement follows standard conventions for the typing judgement of a simply-typed lambda calculus in most respects. The typing context $\Gamma$ maps variables to types and obeys standard structural properties. The key novelties here are 1) that the available type and operator constructors are not fixed by the language, but rather are tracked by the constructor context, $\fvalCtx$, and 2) that a translation is simultaneously derived. There is no separate operational semantics for the external language; the dynamic semantics are defined by this translation to the internal language (which might have a more conventional operational semantics). \todo{cite Harper-Stone / denotational semantics} The two contexts and the external term can be thought of as input to the judgement, while the type and translation are output (we do not consider a bidirectional approach \cite{LovasPfenning}\todo{cite} or type inference in this work, though this is likely a promising avenue for future research).

In our calculus, the internal language (IL), consisting of internal terms $\iota$ and internal types $\sigma$, provides partial functions (via the generic fixpoint operator of Plotkin's PCF \cite{PCF}), simple product and sum types and a base type of integers (to make our example interesting and as a nod toward speed on contemporary machines). In practice, the internal language could be any typed  language with a specification for which type safety and decidability of typechecking have been satisfyingly determined (e.g. Standard ML or CakeML \cite{cakeml}, leaving translation to a simpler IL to later stages of the compilation process).

@@

\inferrule[hide-trans]{
	\fvar{Tycon} \notin \Xi
}{
	\checkRCX{\itransof{\tden{\ibar}{\ttype{Tycon}{\tauidx}}^{\checkmark}}}{\trepof{\ttype{Tycon}{\tauidx}}}
}
\end{mathpar}
\caption{\small Abstracted internal typing}
\label{ait}
\end{figure}
@$\lambda$ is an ``actively typed'' version of the simply-typed lambda calculus with simply-kinded type-level computation. Specifically, type-level terms, $\tau$, themselves form a typed lambda calculus. The classifiers of terms at the type level are called \emph{kinds}, $\kappa$ to distinguish them from  \emph{types}, which are just one kind  of type-level value. We will write the kind of types as $\kTypeBlur$, though it is also written $\star$, \verb|T| or \verb|Type| in various similarly structured languages (see Sec. \ref{related-work}). The kinding rules for type-level terms are given in Fig. \ref{todo}\todo{refer}.% Rather than there being a fixed set of type constructors, we allow the programmer to declare new type  constructors, and give the static and dynamic semantics of their associated operators, by writing type-level functions. In the semantics for this calculus, our kind system combined with techniques borrowed from the typed compilation literature and a form of type abstraction allow us to prove strong type safety, decidability and conservativity theorems.

In most languages, types are formed by applying one of a fixed collection of \emph{type constructors} to zero or more \emph{indices}. In @$\lambda$, user-defined type constructors can be declared at the top-level of a program (or lifted there, in a practical implementation) using \textsf{tycon} (Fig. \ref{kindprog}). Each type constructor in the program must have a unique name, written e.g. \fvar{Nat}.\footnote{We assume naming conflicts can be avoided by some extrinsic mechanism.} A type constructor must also declare an \emph{index kind}, $\kappaidx$. Introducing a type is thus type constructor application to an index, written $\ttype{Tycon}{\tauidx}$. The kind $\kTypeBlur$ also has an elimination form: a type can be case analyzed against an available type constructor to extract its index. We are, to a first approximation, treating type constructors as constructors of a built-in open datatype at the type-level. Like open datatypes, there is no longer a notion of exhaustiveness and the default case is required for case analysis.

 To permit the implementation of interesting type systems, the type-level language includes several kinds other than $\kTypeBlur$. We lift several  standard functional fragments to the type level: unit ($\kunit$), binary sums ($\ksum{\kappa_1}{\kappa_2}$), binary products ($\kpair{\kappa_1}{\kappa_2}$), lists ($\klist{\kappa}$) and integers ($\dint$). We also include labels ($\klabel$), written in a slanted font, e.g. $\tlabel{myLabel}$, which are string-like values that only admit comparisons, here only for equality, and play a distinguished role in the expanded syntax, as we will later discuss (strings themselves could also be included, but we omit them for concision). Our first example, $\fvar{Nat}$, is indexed trivially, i.e. by unit kind, $\kunit$, because there is only one natural number type, $\ttype{Nat}{\tunit}$, but we will show examples of type constructors that are indexed in more interesting ways in later portions of this work. For example, $\fvar{Tuple}$ has index kind $\klist{\kTypeBlur}$ and $\fvar{LabeledTuple}$ has index kind $\klist{\kpair{\klabel}{\kTypeBlur}}$. The type constructor $\fvar{Arrow}$ is included in the initial constructor context and has index kind $\kpair{\kTypeBlur}{\kTypeBlur}$.

Two type-level terms of kind $\kTypeBlur$ are equivalent if they apply the same constructor, identified by name, to equivalent indices. Going further, we ensure that deciding type equivalence requires only checking for syntactic equality after normalization by imposing the restriction that type constructors can only be indexed by kinds for which equivalence can be decided in this way. All combinations of kinds introduced thus far  have this property and this is sufficient for many interesting examples. Our treatment of equivalence in the type-level language is thus quite similar to the treatment of term-level equality using ``equality types'' in a language like Standard ML. An \emph{equality kind} is defined by the judgement \todo{eq type discussion}. We discuss introducing a richer notion of  equivalence of types that  fragment providers can extend as future work in Sec. \ref{limitations}\todo{do this}.% as maintaining the metatheoretic guarantee that typing respects type equivalence would impose a substantial burden in such a setting.% (a na\"ive approach to this would impose non-trivial extrinsic proof obligations onto extension developers that, unlike in others in this thesis, could threaten type safety).

Type constructors are not first-class; they do not themselves have arrow kind as in some kind systems (e.g.  \cite{Watkins08}\todo{cite}; Ch. 22 of \emph{PFPL} describes a related system \cite{pfpl}). The type-level language does, however, include total functions of conventional arrow kind, written $\karrow{\kappa_1}{\kappa_2}$. Type constructor application can be wrapped in a type-level function to emulate a first-class type constructor (indeed, such a wrapper could be generated automatically, though we avoid this for simplicity in our semantics). Equivalence at arrow kind does not meet our criteria, so type-level functions cannot appear within type indices. This also prevents general recursion from arising at the type level. Without this restriction, a function of type $\karrow{\kTypeBlur}{\kappa}$ could ``smuggle'' in a self-reference as a type index and extract it via case analysis (an issue related to the positivity condition for inductive datatypes in total functional languages).

Every type constructor also defines a \emph{representation schema} using \textsf{schema}, a type-level function that we will discuss in Sec. \ref{?}\todo{ref} after introducing operator constructors.

\subsection{Operator Constructors}
For reasons that we will discuss, our calculus associates every operator  constructor\footnote{It may be helpful to emphasize that \emph{term formers} are distinct from type constructors and operator constructors. There are term formers at all levels in the calculus. For example, type constructor application is a type-level term former. An abstract syntax at the type level like Harper's \cite{pfpl} might write it $\texttt{tcapp}[\fvar{Tycon}](\tau)$, emphasizing the distinction.} with a type constructor.

User-defined operator constructors are declared using \textsf{opcon} (Fig. \ref{kindprog}). The \emph{fully-qualified name} of every operator constructor, e.g. $\fvar{Nat}.\opvar{z}$, must be unique. Operator constructors are indexed by type-level values, like type constructors, and so also declare an index kind $\kappaidx$. In our first example, all the operator constructors are indexed trivially (by index kind $\kunit$), but later examples will use more interesting indices. For example, the projection operator for labeled tuples, $\fvar{LabeledTuple}.\opvar{prj}$, has index kind $\klabel$. \todo{defne operator}
Note that neither operator constructors nor operators are first-class type-level values and we do not need to impose any restrictions on their index kinds.

In the external language, an operator is selected and invoked by applying an operator constructor to a type-level index and $n \geq 0$ \emph{arguments}, written $\eop{Tycon}{op}{\tauidx}{\splat{e}{1}{n}}$. For example, on line 18 of Fig. \ref{nat}, we see the operator constructors $\fvar{Nat}.\opvar{z}$ and $\fvar{Nat}.\opvar{s}$ being invoked to compute $two$ (the syntactic overhead will be relieved later). %\footnote{Although our focus here is entirely on semantics, a brief note on syntax: in the expanded syntax, the trivial indices and empty argument lists can be omitted, so we could write \texttt{Nat.s(Nat.s(Nat.z))}. With the ability to ``open'' a type's operators into the context, we could shorten this still to \texttt{s(s(z))}. Alternatively, with the ability to define a TSL in a manner similar to that in Sec. \ref{aparsing}, we might instead just write \texttt{2}.}
To derive the active typing judgement for this form, the semantics invokes  
 the operator constructor's  \emph{definition}: a type-level function that must examine the provided operator index and the recursively determined \emph{derivates} of the arguments to determine a derivate for the operation as a whole,  or indicate an error. A derivate is a type-level representation of the result of deriving the active typing judgement\footnote{Technically, the abstract active typing judgement, which is similar in form and we will introduce shortly.}: an internal term, called the {translation},  paired with a type. Derivates have kind $\kDen$ and introductory form $\tden{\tautrans}{\tautype}$. The elimination form $\tderbind{x}{t}{\tau}{\tau'}$ extracts the translation and type from the derivate $\tau$, binding it to $\tvar{x}$ and $\tvar{t}$ respectively in $\tau'$. 
%Indeed, one might be tempted to simply define derivates as products. However, we treat translations with much care throughout the calculus, so this is not quite the case (we will return to this shortly). 
Because constructing a derivate is not always possible (e.g. when there is a type error in the client's code, or an invalid index was provided), the return kind of the function is an ``option kind'', $\ksum{\kDen}{\kunit}$, where the trivial case indicates a statically-detected error in the client's program. In practice, it would instead require providers to report information about the precise location of the error and an appropriate error message.
